---
title: "Class 16"
author: "Joseph Girgiss (PID: A17388247)"
format: pdf
toc: true
---

## Plot my results

```{r}
library(readr)
b <- read_tsv("my_results.tsv", col_names = c("qseqid", "sseqid", "pident", "length", "mismatch", "gapopen", "qstart", "qend", "sstart", "send", "evalue", "bitscore"))
```

```{r}
library(ggplot2)

ggplot(b, aes((b$pident * (b$qend - b$qstart)), bitscore)) + 
  geom_point(alpha=0.1) + 
  geom_smooth()
```


## Lab 16 Questions for Gradescope

> Q1. [6pt] List the UNIX bash shell commands to: 

1.1 open a secure shell on a remote machine: ssh josephgirgiss@biglabcluster.ucsd.edu

1.2 make a new folder in your home area called "test": mkdir ~/test

1.3 download this file "https://files.rcsb.org/download/5P21.pdb.gz": curl -0 https://files.rcsb.org/download/5P21.pdb.gz

1.4unzip/decompress it the file: gunzip 5P21.pdb.gz

1.5 print to screen the first 6 lines: head -6 5P21.pdb 

1.6 print to lines beginning with ATOM to a new file called "coords.pdb": grep "^ATOM" 5P21.pdb > coords.pdb

>Q2. [3pt] List the UNIX commands to copy securely the file "myaln.fa" in your current working directory to your home area on the remote machine "biglabcluster.ucsd.edu":

scp myaln.fa josephgirgiss@biglabcluster.ucsd.edu:~/.

> Q3. [1pt] The alignment file “myaln.fa” is not in your current working directory but it is in your “Downloads” directory. Write the R code to import this alignment to the named object “aln” using a function from the bio3d package. 

library(bio3d)
aln <- read.fasta("~/Downloads/myaln.fa")


