---
title: "Class 17"
author: "Joseph Girgiss (PID: A17388247)"
format: pdf
---

## Section 1: Proportion of G|G in a population

Downloaded a CSV file from Ensembl. 

Here we read this CSV file 

```{r}
mxl <- read.csv("373531-SampleGenotypes-Homo_sapiens_Variation_Sample_rs8067378.csv")
head(mxl)
```


```{r}
table(mxl$Genotype..forward.strand) / nrow(mxl) * 100
```

Now let's look at a different population. I picked the GBR population. 

```{r}
gbr <- read.csv("373522-SampleGenotypes-Homo_sapiens_Variation_Sample_rs8067378 (2).csv")

# find G|G proportion
round(table(gbr$Genotype..forward.strand) / nrow(gbr) * 100, 2)
```
This variant that is associated with childhood asthma is more frequent in the GBR population than the MXL population. 

Let's explore this further. 


## Section 4: Population scale analysis
One sample is obviously not enough to know what is happening in a population. You are
interested in assessing genetic differences on a population scale. So, you processed about
~230 samples and did the normalization on a genome level. Now, you want to find whether
there is any association of the 4 asthma-associated SNPs (rs8067378…) on ORMDL3
expression.

How many samples do we have? Let's read the data. 

> Q13: Read this file into R and determine the sample size for each genotype and their
corresponding median expression levels for each of these genotypes.

```{r}
expr <- read.table("rs8067378_ENSG00000172057.6.txt")
head(expr)
```

```{r}
nrow(expr)

```

```{r}
table(expr$geno)
```
```{r}
tapply(expr$exp, expr$geno, median)
```
Let's make a boxplot

>Q14: Generate a boxplot with a box per genotype, what could you infer from the relative
expression value between A/A and G/G displayed in this plot? Does the SNP effect the
expression of ORMDL3?

```{r}
library(ggplot2)
ggplot(expr) + aes(geno, exp, fill=geno) + 
  geom_boxplot(notch=TRUE)
```

The A/A genotype shows the highest expression of ORMDL3, while A/G and G/G show progressively lower expression levels. This pattern suggests a allelic effect of the SNP on gene expression, where each copy of the G allele is associated with reduced ORMDL3 expression. So, yes, SNP does effect the expression of ORMDL3. 






