---
title: "Class 10: Halloween mini project"
author: "Joseph Girgiss (PID: A17388247)"
format: pdf
toc: TRUE
---

As it is nearly Halloween and the half way point in the quarter let's do a mini project to help us figure out the best candy! 

Our data comes from the 538 website and is available as a CSV file: 

## Data Import

```{r}
candy_file <- "candy-data.csv"
candy = read.csv(candy_file, row.names=1)
head(candy)
```

> Q1. How many different candy types are in this dataset?

```{r}
nrow(candy)
```

> Q2. How many fruity candy types are in the dataset?

```{r}
sum(candy$fruity)
```

```{r}
candy["Twix", ]$winpercent
```

> Q3. What is your favorite candy in the dataset and what is it’s winpercent value?

My favorite candy in the dataset is Milky Way. 

```{r}
candy["Milky Way", ]$winpercent
```

> Q4. What is the winpercent value for “Kit Kat”?

```{r}
candy["Kit Kat", ]$winpercent
```

> Q5. What is the winpercent value for “Tootsie Roll Snack Bars”?

```{r}
candy["Tootsie Roll Snack Bars", ]$winpercent
```


## Quick overview of the dataset

```{r}
library(skimr)
skimr::skim(candy)
```

> Q6. Is there any variable/column that looks to be on a different scale to the majority of the other columns in the dataset?

Yes, winpercent looks to be on a different scale since its values between 0-100 when all others are between 0 and 1. 

> Q7. What do you think a zero and one represent for the candy$chocolate column?

```{r}
candy$chocolate
```

1 = the candy is a chocolate, 0 = the candy is not a chocolate. 


> Q8. Plot a histogram of winpercent values

```{r}
library(ggplot2)

ggplot(candy, aes(winpercent)) +
  geom_histogram(binwidth = 5, fill = "steelblue", color = "white") +
  labs(title = "Distribution of Candy Win Percentages",
       x = "Win Percent", y = "Count") + 
  theme_bw()
```

> Q9. Is the distribution of winpercent values symmetrical?

```{r}
ggplot(candy, aes(winpercent)) +
  geom_density() + 
  theme_bw()
```

The distribution of winpercent values is asymmetrical. 

> Q10. Is the center of the distribution above or below 50%?

```{r}
mean(candy$winpercent)
summary(candy$winpercent)
```
Center: mean = 50.31676 > 50%, median = 47.83 < 50% (indicating a right skew). 

> Q11. On average is chocolate candy higher or lower ranked than fruit candy?

```{r}
# 1. Find all chocolate candy in the dataset 
# 2. Find their winpercent values 
# 3. Calculate the mean of these values 
# 4. Repeat for fruity candy. 
# 5. Compare mean winpercents of chocolate vs fruity

choc.inds <- candy$chocolate==1
choc.win <- candy[choc.inds, ]$winpercent
choc.mean <- mean(choc.win)
choc.mean

fruit.inds <- candy$fruity==1
fruit.win <- candy[fruit.inds, ]$winpercent
fruit.mean <- mean(fruit.win)
fruit.mean

```

```{r}
library(dplyr)
candy %>%
  summarize(
    mean_choc = mean(winpercent[chocolate == 1]),
    mean_fruit = mean(winpercent[fruity == 1])
  )
```
On average the chocolate candy is higher than fruity candy. 

> Q12. Is this difference statistically significant?

```{r}
t.test(choc.win, fruit.win)
```

The difference between chocolate and fruity candy is statistically significant because the p-value is very small (p-value = 2.871e-08).


> Q13. What are the five least liked candy types in this set?

```{r}
candy %>% 
  arrange(winpercent) %>%
  head(5)
```

> Q14. What are the top 5 all time favorite candy types out of this set?

```{r}
candy %>% 
  arrange(desc(winpercent)) %>%
  head(5)
```

> Q15. Make a first barplot of candy ranking based on winpercent values.


```{r}
library(ggplot2)
candy$CandyName <- factor(rownames(candy), levels = rownames(candy))


ggplot(candy) +
  aes(x = winpercent, y = CandyName) +
  geom_col() +   
  theme_bw() +
  labs(x = "Win Percent", y = "Candy", title = "Candy Rankings by Win Percent") +
  theme(
    axis.text.y = element_text(size = 4)   
  )
```

> Q16. This is quite ugly, use the reorder() function to get the bars sorted by winpercent?

Let's also add some color based on the type of candy. 

```{r}
my_cols=rep("black", nrow(candy))
my_cols[as.logical(candy$chocolate)] = "chocolate"
my_cols[as.logical(candy$bar)] = "brown"
my_cols[as.logical(candy$fruity)] = "pink"
```

```{r}
candy$CandyName <- factor(rownames(candy), levels = rownames(candy)[order(candy$winpercent)])
ggplot(candy) +
  aes(x = winpercent, y = CandyName) +
  geom_col(fill=my_cols) +   
  theme_bw() +
  labs(x = "Win Percent", y = "Candy", title = "Candy Rankings by Win Percent") +
  theme(
    axis.text.y = element_text(size = 4)   
  )
```

> Q17. What is the worst ranked chocolate candy?

```{r}
candy %>% 
  filter(chocolate == 1) %>%
  arrange(winpercent) %>%
  head(1)
```
Sixlets are the worst ranked chocolate candy. 

> Q18. What is the best ranked fruity candy?

```{r}
candy %>% 
  filter(fruity == 1) %>%
  arrange(desc(winpercent)) %>%
  head(1)
```
 Starburst is the best ranked fruity candy. 

## Winpercent and Pricepercent

A plot with both variable/columns winpercent and pricepercent

```{r}
library(ggrepel)

# How about a plot of price vs win
ggplot(candy) +
  aes(winpercent, pricepercent, label=rownames(candy)) +
  geom_point(col=my_cols) + 
  geom_text_repel(col=my_cols, size=3.3, max.overlaps = 5)
```


> Q19. Which candy type is the highest ranked in terms of winpercent for the least money - i.e. offers the most bang for your buck?

```{r}
library(dplyr)
candy %>%
  mutate(Value = winpercent / (pricepercent + 1e-6)) %>%
  arrange(desc(Value)) %>%
  head(1)
```


> Q20. What are the top 5 most expensive candy types in the dataset and of these which is the least popular?

```{r}
top5_expensive <- candy %>%
  arrange(desc(pricepercent)) %>%   # sort by price descending
  slice(1:5) %>%                    # take top 5
  select(pricepercent, winpercent)  # show relevant columns

top5_expensive

least_popular <- top5_expensive %>%
  slice_min(winpercent)   # candy with lowest winpercent

least_popular
```


## Exploring the correlation structure

Now that we’ve explored the dataset a little, we’ll see how the variables interact with one another. We’ll use correlation and view the results with the corrplot package to plot a correlation matrix.

```{r}
library(corrplot)
cij <- cor(candy[sapply(candy, is.numeric)])
corrplot(cij)
```

> Q22. Examining this plot what two variables are anti-correlated (i.e. have minus values)?

```{r}
# Convert the correlation matrix to a data frame for easier filtering
corr_df <- as.data.frame(as.table(cij))
colnames(corr_df) <- c("Variable1", "Variable2", "Correlation")

# Filter out the diagonal (where Variable1 equals Variable2)
off_diagonal_corr <- subset(corr_df, Variable1 != Variable2)

# Find the row with the minimum correlation value
most_anti_correlated <- off_diagonal_corr[which.min(off_diagonal_corr$Correlation), ]

# Print the result
print("Most Anti-Correlated Pair:")
print(most_anti_correlated)
```
Fruity and Chocolate are most anti-correlated. 

> Q23. Similarly, what two variables are most positively correlated?

```{r}
# Find the row with the maximum correlation value
most_positively_correlated <- off_diagonal_corr[which.max(off_diagonal_corr$Correlation), ]

# Print the result
print("Most Positively Correlated Pair:")
print(most_positively_correlated)
```

Winpercent and chocolate are the most potively correlated pair. 

## Principal component analysis

The function to use is called `prcomp()` with an optional `scale=T/F` argument. 

```{r}
candy_numeric <- candy[sapply(candy, is.numeric)]
pca <- prcomp(candy_numeric, scale=TRUE)
summary(pca)
```

Our main PCA result figure

```{r}
ggplot(pca$x) + 
  aes(PC1, PC2) + 
  geom_point(col=my_cols) +
  geom_text_repel(col=my_cols,label=rownames(candy)) + 
  theme_bw()
```

We should also examine the variable "loadings" or contributions of the original variable to the new PCs. 

```{r}
pca$rotation
ggplot(pca$rotation) + 
  aes(PC1, rownames(pca$rotation)) +
  geom_col()
```

```{r}
plot(pca$x[,1:2], col=my_cols, pch=16)
```


```{r}
# Make a new data-frame with our PCA results and candy data
my_data <- cbind(candy, pca$x[,1:3])

p <- ggplot(my_data) + 
        aes(x=PC1, y=PC2, 
            size=winpercent/100,  
            text=rownames(my_data),
            label=rownames(my_data)) +
        geom_point(col=my_cols)

p
```

```{r}
library(ggrepel)

p + geom_text_repel(size=3.3, col=my_cols, max.overlaps = 7)  + 
  theme(legend.position = "none") +
  labs(title="Halloween Candy PCA Space",
       subtitle="Colored by type: chocolate bar (dark brown), chocolate other (light brown), fruity (red), other (black)",
       caption="Data from 538")
```
Interactive plots that can be zoomed on and "brushed" over can be made with the **plotly** package. It's output is interactive and will not render to PDF. 

```{r}
#library(plotly)
#ggplotly(p)
```


```{r}
par(mar=c(8,4,2,2))
barplot(pca$rotation[,1], las=2, ylab="PC1 Contribution")
```

> Q24. What original variables are picked up strongly by PC1 in the positive direction? Do these make sense to you?

Fruity, hard, and pluribus are picked up strongly by PC1 in the positive direction. 
This makes sense because PC1 can be considered as a comparison of Chocolate vs fruity. It is logical that the features defining a typical non-chocolate/hard/many-pieces candy (Fruity, Pluribus, Hard) are strongly aligned in the positive direction of PC1, contrasting with the chocolate/bar/high-value (high pricepercent and winpercent) features in the negative direction.




